Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-15 text/manuscript-full-KS.qmd
worked through quantification section on full manuscript

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of EM and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Scholarship from the economics of education often investigate financial aid leveraging, which seeks to convert admits to enrolled students [e.g., @RN3762; @RN3133]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

By focusing on the behavior of colleges, scholarship on EM implicitly assumes that recruiting is something done by individual colleges. The broader EM industry includes relevant stakeholders in the organizational field [@RN527], including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways EM contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. Unfortunately, few empirical studies make EM vendors or products the object of empirical analysis. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

[^2]: In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from a student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

Quantification can be defined as "the production and communication of numbers" [@RN5007, p. 402]. While a growing interdisciplinary area of scholarship, sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, and personal life [@RN5008]. Quantification has also been at the core of recent critical data scholarship exploring how algorithms and big data computation structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008 (p. 239), to explore the role of third-party servicers and products in the sorting of students into colleges. Specifically, we draw on quantification scholarship investigating how correlation and homophily [@RN4975] encode social reproduction into our growingly dominant "algorithmic society" \[[CITE](https://medium.com/chatbots-bible-and-more/theres-an-algorithm-for-that-or-there-soon-will-be-d54406dc9bc1)\] and explore how such mechanisms are applied to the construction of Geomarkets in the Market Segment Model [@RN4982].

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that

> correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'

Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with an variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Many studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point in time. Observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^3] As we discuss below, @RN4982 analyzed 1980 score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities that have a high share of affluent, college-educated households.

[^3]: An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. Note that predictive models such as the LSI-R model are not merely used for social science purposes. Rather, they reproduce structural inequality because they direct the allocation of future resources based on a snapshot of historical inequality.

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science, in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). In *The company We Keep: Colleges and Their Competition*, @RN4982 [chapter 4] conduct a network analysis whereby two colleges are defined as competitors if a large number of students submit SAT scores to both colleges. @RN4975 argues that because network science models often draw from rational choice theory, they assume that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar phenomena in our case. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

[^4]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.

We make targeted contributions to scholarship on quantification in the sociology of education. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of EM focus on the behavior of colleges and their agents, yielding the implicit belief that inequalities created by EM are a function of individual colleges reacting to macro structures in their environment such as rankings and financial pressures. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of producers. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027]. Drawing from critical data studies [e.g., @RN4775; @RN4975; @RN4849], recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches vertically categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociologists. We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.

<!--# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. By showing which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

[^5]: For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.

[^6]: Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.[^7] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

[^7]: This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."

@@

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^8] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

[^8]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

@@

Quantification disciplines actors to react in particular ways. EPS software may "discipline" [@RN5007] colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.[^9] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

[^9]: Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."

>>>>

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional marketing funnel to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of EM and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Scholarship from the economics of education often investigate financial aid leveraging, which seeks to convert admits to enrolled students [e.g., @RN3133; @RN3762]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

By focusing on the behavior of colleges, scholarship on EM implicitly assumes that recruiting is something done by individual colleges. The EM industry includes relevant stakeholders in the organizational field [@RN527], including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers that supply products and consulting solutions to colleges (e.g., College Board, EAB). We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways the broader EM industry contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].^[In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].]

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The College Board Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The College Board Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from a student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

@RN5007[p. 402] define quantification as "the production and communication of numbers." The effects of quantification are described by the three themes of reactivity (people and orginizations change behavior in response to salient quantitative measures), discipline (quantification disciplines actors to react in particular ways), and authority (quantification often weakens discretion of local decision-makers). Sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, personal life and education, with a particular emphasis on rankings [@RN5008; @RN5003; @RN5005]. In the field of critical data studies, quantification is central to scholarship that explores how computation and algorithms structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008[p. 239], to explore the role of third-party servicers and products in the sorting of students into colleges. In particular, the discussions of correlation and homophily by @RN4975 show how social reproduction is encoded in algorithms that direct future resources. We explore how correlation and homophily were incorporated into the construction and application of Geomarkets within the Market Segment Model [@RN4982].

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

Due to data limitations, predictive analytics often utilize "proxy" variables -- variables highly correlated with a variable of interest [@RN4849; @RN4975]. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access [e.g., @RN4565], when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Many studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point in time. Observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].^[An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests.] As we discuss below, @RN4982 analyzed 1980 SAT score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities with many affluent, college-educated households.

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science, in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). In *The company We Keep: Colleges and Their Competition*, @RN4982 [chapter 4] conduct a network analysis whereby two colleges are defined as competitors if a large number of students submit SAT scores to both colleges. @RN4975 observes that network science models often draw from rational choice theory, which leads to the assumption that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. 94]. We observe similar phenomena in our case. Based on a network analysis of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].^[For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit.] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

We make targeted contributions to scholarship on quantification in the sociology of education. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of EM focus on the behavior of colleges and their agents, creating the sense that inequalities created by EM are a function of individual colleges reacting to macro structures in their environment such as rankings and financial pressures [e.g., @RN5003; @RN3519]. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of producers. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027]. Recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches vertically categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociology.

<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->

This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.^[This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.^[@RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

@@

Quantification disciplines actors to react in particular ways. EPS software may "discipline" [@RN5007] colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.
